---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

# Reproducible Research: Peer Assessment 1
## by Francisco Marin, Coursera 2015 

## </br>
## Loading and preprocessing the data
The first step is to load the file containing the data source.
```{r echo=TRUE}
rawds <- read.csv("activity.csv", head=TRUE, sep=",", col.names = c("steps","date","interval"))
```
The second step is to build a new column for the date, using the correct object type. 
The format was specified in the definition of the problem ("YYYY-MM-DD").
```{r echo=TRUE}
library(dplyr)
rawds <- mutate(rawds, fdate = as.Date(date, "%Y-%m-%d"))
```

## </br>
## What is mean total number of steps taken per day?
It is required to calculate the total of steps by day, the mean and median. It is important to mention 
that the NA values will be ignored.
```{r echo=TRUE}
dategrp <- group_by(rawds, fdate)
stepsds <- summarise(dategrp, TotalSteps = sum(steps, na.rm = TRUE), StepsMean = mean(steps, na.rm = TRUE), 
StepsMedian = median(steps, na.rm = TRUE))
```
The total of steps can be represented in a histogram:
```{r echo=TRUE, fig.width=6, fig.height=6}
hist(stepsds$TotalSteps, main="Histogram of Total Steps by day", xlab="Total Steps by day", col = "lightblue")
```

The mean and median for the total steps in each day are:
```{r kable, echo=TRUE}
kable(stepsds, digits = 2, format="html", longtable = TRUE)
```

## </br>
## What is the average daily activity pattern?
In order to display a pattern of the activity it is necessary to pre-calculate another dataset.
```{r echo=TRUE}
patterngrp <- group_by(rawds, interval)
patternds <- summarise(patterngrp, TotalSteps = sum(steps, na.rm = TRUE), 
	IntervalMean = mean(steps, na.rm = TRUE), IntervalMedia = median(steps, na.rm = TRUE))
```

With the new data set, the pattern of the intervals for the steps can be visualize.
```{r echo=TRUE, fig.width=6, fig.height=6}
plot(patternds$interval, patternds$IntervalMean, type="l", col = "blue", main="Mean of steps by Interval", 
	xlab="Interval", ylab="Mean of steps")
```

There is a clear peak in the graph. It is possible to get the exact interval with the maximum average steps:
```{r echo=TRUE}
patternds[patternds$IntervalMean == max(patternds$IntervalMean), c(1,3)]
```


## </br>
## Imputing missing values
First, calculate the number of NA values in the original data set.
```{r echo=TRUE}
numNA <- length(rawds[is.na(rawds$steps),3])
numNA
```

In this section the script will fill the NA values. The selected strategy is to use the median
for the interval. 
```{r echo=TRUE}
simds <- rawds
stepsNum <- length(rawds$steps)
for(idx in 1:stepsNum) {
	if(is.na(simds$steps[idx])) {
		simds$steps[idx] <- patternds[ patternds$interval == simds$interval[idx] , 4]
	}
}

# The previous step will change the data type to list, the next step
# will convert it back to numeric
simds$steps <- as.numeric(unlist(simds$steps))
```

After filling the NA data, the next step is to recreate the pre-calculated data frame:
```{r echo=TRUE}
dategrpNA <- group_by(simds, fdate)
stepsdsNA <- summarise(dategrpNA, TotalSteps = sum(steps), StepsMean = mean(steps), 
StepsMedian = median(steps))
```

With the calculated data, it is possible to generate the 
```{r echo=TRUE, fig.width=6, fig.height=6}
hist(stepsdsNA$TotalSteps, main="Histogram of Total Steps by day with Corrected NA values",
	xlab="Total Steps by day", col = "lightblue")
```

The median and median for each day can be visualize as:
```{r echo=TRUE}
kable(stepsdsNA, digits = 2, format="html", longtable = TRUE)
```

After the corrections the distribution of the means has the same shape in the histograms. 
However, the Total Number of Steps has a major difference. A good example is comparing
the sum of the steps for each data set:
```{r echo=TRUE}
totalStepsWithoutNA <- sum(stepsdsNA$TotalSteps)
totalStepsWithNA <- sum(stepsds$TotalSteps)
paste("Sum of steps in original data frame:", totalStepsWithNA, sep=" ")
paste("Sum of steps in modified data frame:", totalStepsWithoutNA, sep=" ")
```

## </br>
## Are there differences in activity patterns between weekdays and weekends?

The first step is to create a new factor variable with two levels: weekend and weekday.
```{r echo=TRUE} 
simds$daytype <- ifelse(weekdays(simds$fdate) %in% c("Sunday","Saturday"), "weekend","weekday")
```

After creating the categorical data, a new data frame must be created to store the averages of steps per interval.
```{r echo=TRUE}
patterngrpdaytyp <- group_by(simds, daytype,interval)
patterndsdaytyp <- summarise(patterngrpdaytyp, MeanSteps = mean(steps))
```

With the new categorical variable it is possibly to see the difference between weekdays and weekends:
```{r echo=TRUE} 
par(mfrow = c(2,1))
weekends <- patterndsdaytyp[patterndsdaytyp$daytype == "weekend", ]
weekdays <- patterndsdaytyp[patterndsdaytyp$daytype == "weekday", ]
plot(weekends$interval, weekends$MeanSteps, type="l", col = "blue", 
	main="Mean of steps by Interval in Weekends", 
	xlab="Interval", ylab="Mean of steps")
plot(weekdays$interval, weekdays$MeanSteps, type="l", col = "blue", 
	main="Mean of steps by Interval in Weekdays", 
	xlab="Interval", ylab="Mean of steps")
```